RationalGRL: A Framework for Argumentation and Goal Modeling

### 2017-08-22 introduction.tex
Major update section 1, some changes section 2.4, references added

\subsection{Requirements Engineering and Goal Modeling}
\todo{Floris}{all}{Starting section 1 directly with subsection 1.1 is generally frowned upon. Maybe we can get rid of the subsections in the intro?} 
Requirements Engineering (RE) is an approach to assess the role of a future information system within a human or automated environment. An important goal in RE is to produce a consistent and comprehensive set of requirements covering different aspects of the system, such as general functional requirements, operational environment constraints, and so-called non-functional requirements such as security and performance.

@@

\subsection{Problem Statement}
\label{sect:problem_statement}

A goal model is often the result of a discussion process between a group of stakeholders. For small-sized systems, goal models are usually constructed in a short amount of time, involving stakeholders with a similar background. Therefore, it is often not necessary to record all of the details of the discussion process that led to the final goal model. However, goal models for many complex, real-world information systems -- e.g., air-traffic management systems, systems that support industrial production processes, or government and healthcare services -- are not constructed in a short amount of time, but rather over the course of several workshops with stakeholders and requirements engineers. In such situations, failing to record the discussions underlying a goal model in a structured manner may harm the success of the RE phase of system development for various reasons.

First, it is well-known that stakeholders' preferences are rarely absolute, relevant, stable, or consistent~\cite{march1978bounded}. Therefore, it is possible that a stakeholder changes his or her opinion about a modeling decision in between two goal modeling sessions, which may require revisions of the goal model. If previous preferences and opinions are not stored explicitly, it is not straightforward to remind stakeholders of their previous opinions which can result in unnecessary discussions and revisions. As the number of participants increases, revising the goal model based on changing preferences can take up a significant amount of time.

Second, other stakeholders, such as new developers on the team who were not the original authors of the goal model, may have to make sense of the goal model, for instance, to use it as an input in a later RE stage or at the design and development phase. If these users have no knowledge of the underlying rationale of the goal model, it may not only be more difficult to understand the model, but they may also end up having the same discussions as the previous group of stakeholders.

Third, alternative and different ideas and opposing views that could potentially have led to different goal models could be lost. For instance, a group of stakeholders specifying a goal model for a user interface may decide to reduce the two goals ``easy to use'' and ``fast'' to one goal ``easy to use''. Thus, the resulting goal model merely contains the goal ``easy to use'', but the discussion as well as the decision to reject the goal ``fast'' are lost. This leads to a poor understanding of the problem and solution domain. In fact, empirical data suggest that this is an important reason of RE project failure~\cite{curtis1988field}.

Fourth and finally, in goal models in general the rationale behind any modeling decision is static and changing these rationales does not immediately impact the goal models. That is, current goal modeling languages have limited support for reasoning about changing beliefs and opinions, and their effect on the goal model. A stakeholder may change his or her opinion, but it is not always directly clear what its effect is on the goal model. Similarly, with existing goal modelling languages one can change a part of the goal model, but it is not possible to reason about whether or not this new goal model is consistent with the underlying beliefs and arguments. This becomes more problematic if the participants constructing the goal model change, since modeling decisions made by one group of stakeholders may conflict with the underlying beliefs put forward by another group of stakeholders.

\todo{Floris}{Floris}{I think the above 4 reasons overlap in some cases. If it's OK, I'll rewrite this.}

\subsection{Research Questions}

The aim of this research is to resolve the above issues by developing a framework with tool-support, which combines goal modeling approaches with argumentation techniques from Artificial Intelligence (AI) research~\cite{atkinson2007}. We have identified five important requirements for our framework: 
\begin{enumerate}
\item The argumentation techniques should be close to the actual discussions of stakeholders or designers in the early requirements engineering phase.
\item 
The framework must have formal traceability links between elements of the goal model and underlying arguments.
\item 
Using these traceability links, it must be possible to compute the effect of changes in the goal model on the underlying arguments, and vice versa.
\item 
The framework must have software tool support.
\item 
There should be a methodology for the framework to guide the practitioners in its application in real cases.
\end{enumerate}

The five requirements above serve as the success criteria of our approach. That is, if our framework satisfies these requirements, we reach our goal.

In this context, the main research question is:

\begin{quote}
\textbf{RQ.} What are the constructs, mechanisms and rules for developing a framework that formally captures the discussions between stakeholders such that it can generate goal models?
\end{quote}
\todo{Floris}{Marc,Sepideh}{While I like an explicit research question, you introduce new terminology here ``construct, mechanisms and rules''. Do you explicitly refer to this later in the paper (e.g. we found mechanisms A, B, rules R1, R2, etc.)?}
To answer the research question above and solve some of the concerns mentioned in Subsection~\ref{sect:problem_statement}, we propose a framework called RationalGRL. The RationalGRL framework combines the Goal-oriented Requirements Language (GRL) with a technique from argumentation theory and discourse modeling called \emph{argument schemes}~\cite{walton-etal2004}. \todo{Floris}{Floris}{mention Singh et al. argument schemes in RE here.}

In previous work, we have taken preliminary steps towards the RationalGRL framework~\cite{bnaic2014,vanzee-etal:renext2015,vanZee-etal:er2016} with its tool-support~\cite{vanZee-etal:comma2016}.  In our previous work, argument diagrams are integrated with GRL models to allow better representation of stakeholders' arguments and discussions. In our current paper, we improve the RationalGRL framework by adding argument schemes and critical questions to argumentation framework and provide traceability links between the argumentation framework and GRL.  For this, we modify and extend our metamodel to specify how our framework extends GRL, and we use this as the basis for a prototype web-based implementation\footnote{The tool is available at <GITHUB LINK>}. %SG: Placeholder for the link. 
Inspired by work on practical reasoning in Artificial Intelligence~\cite{atkinson2007}, we develop an initial list of argument schemes and critical questions for reasoning about goals in RE. We do this by analyzing a set of transcripts containing discussions about a traffic simulator information system. Our framework is fully extensible, meaning that argument schemes and critical questions can be added by users for specific problem domains. We develop a methodology for using RationalGRL, which consists of developing goal models and posing arguments in an integrated way. This methodology is intended to be used by practitioners in the field. We provide formal semantics for RationalGRL by translating the elements of the GRL language into simple propositional logical formulas and subsequently capturing arguments about these elements of a GRL model in a formal argumentation framework \cite{dung1995}. We then develop a set of algorithms for applying argument schemes and critical questions about goal models. These algorithms further formally ground the RationalGRL methodology and form the basis of the RationalGRL tool. We illustrate the steps of our framework and how RationalGRL can be applied in a real case with the traffic simulator case study.

\subsection{Organization} %SG: Will get back to this at the end of the paper.

This article is organized as follows. Section~\ref{sect:background} introduces our running example, the Goal-oriented Requirements Language (GRL)~\cite{Amyot:2010:EGM:1841349.1841356}, the Practical Reasoning Argument Scheme (PRAS)~\cite{atkinson2007} and some general background about . Section~\ref{sect:overview} provides a brief and high-level overview of our framework, together with a metamodel and the methodology. Section~\ref{sect:gmas} contains an in depth explanation of how we obtained an initial set of argument schemes and critical questions by annotating transcripts from discussions about an information system, and in Section~\ref{sect:examples} we provide several examples of these schemes and questions. In Section~\ref{sect:formalframework} we provide formal semantics for GRL and show how argumentation semantics \cite{dung1995} can be used to compute which arguments are accepted and which are rejected. We also develop various algorithms for the argument schemes in this section. In Section~\ref{sect:tool} we provide a brief overview of the prototype tool we developed for RationalGRL. Finally, Section~\ref{sect:discussion} contains a discussion, covering related work, future work, and a conclusion.

>>>>

Requirements Engineering (RE) is an approach to assess the role of a future information system within a human or automated environment. An important goal in RE is to produce a consistent and comprehensive set of requirements covering different aspects of the system, such as general functional requirements, operational environment constraints, and so-called non-functional requirements such as security and performance.

@@

A goal model is often the result of a discussion process between a group of stakeholders. For small-sized systems, goal models are usually constructed in a short amount of time, involving stakeholders with a similar background. Therefore, it is often not necessary to record all of the details of the discussion process that led to the final goal model. However, goal models for many complex, real-world information systems -- e.g., air-traffic management systems, systems that support industrial production processes, or government and healthcare services -- are not constructed in a short amount of time, but rather over the course of several workshops with stakeholders and requirements engineers. In such situations, failing to record the discussions underlying a goal model in a structured manner may harm the success of the RE phase of system development.

The first difficulty is that the goal modeling phase, particularly in large projects, is dynamic and that goal models continuously change and evolve. Stakeholders' preferences are rarely absolute, relevant, stable, or consistent~\cite{march1978bounded}, and stakeholders may change their opinion about a modeling decision in between two goal modeling sessions, which in turn may require revisions of the goal model. If the rationales behind these revisions are not properly documented, alternative ideas and opposing views that could potentially have led to different goal models are lost, as the resulting goal model only shows the end product of a long process and not the discussions during the modeling process. Furthermore, other stakeholders, such as developers who were not the original authors of the goal model, may have to make sense of a goal model in order to, for example, use it as input in a later RE stage or in the development phase. If preferences, opinions and rationales behind the goal models are not stored explicitly, it may not only be more difficult to understand the model, but the stakeholders may also end up having the same unnecessary discussions throughout the goal modeling phase.

A further problem is that in goal models in general the rationale behind any modeling decision is static, that is, current goal modeling languages have limited support for reasoning about changing beliefs and opinions, and their effect on the goal model. A stakeholder may change his or her opinion, but it is not always directly clear what its effect is on the goal model. Similarly, with existing goal modelling languages one can change a part of the goal model, but it is not possible to reason about whether or not this new goal model is consistent with the underlying beliefs and arguments. This becomes even more problematic if the stakeholders constructing the goal model change, since modeling decisions made by one group of stakeholders may conflict with the underlying beliefs of another group of stakeholders. The disconnect between the goal models and their underlying beliefs and opinions may further lead to a poor understanding of the problem and solution domain, which is an important reason of RE project failure~\cite{curtis1988field}.

%First, it is well-known that stakeholders' preferences are rarely absolute, relevant, stable, or consistent~\cite{march1978bounded}. Therefore, it is possible that a stakeholder changes his or her opinion about a modeling decision in between two goal modeling sessions, which may require revisions of the goal model. If previous preferences and opinions are not stored explicitly, it is not straightforward to remind stakeholders of their previous opinions which can result in unnecessary discussions and revisions. As the number of participants increases, revising the goal model based on changing preferences can take up a significant amount of time.

%Second, other stakeholders, such as new developers on the team who were not the original authors of the goal model, may have to make sense of the goal model, for instance, to use it as an input in a later RE stage or at the design and development phase. If these users have no knowledge of the underlying rationale of the goal model, it may not only be more difficult to understand the model, but they may also end up having the same discussions as the previous group of stakeholders.

%Third, alternative and different ideas and opposing views that could potentially have led to different goal models could be lost. For instance, a group of stakeholders specifying a goal model for a user interface may decide to reduce the two goals ``easy to use'' and ``fast'' to one goal ``easy to use''. Thus, the resulting goal model merely contains the goal ``easy to use'', but the discussion as well as the decision to reject the goal ``fast'' are lost. This leads to a poor understanding of the problem and solution domain. In fact, empirical data suggest that this is an important reason of RE project failure~\cite{curtis1988field}.

%Fourth and finally, in goal models in general the rationale behind any modeling decision is static and changing these rationales does not immediately impact the goal models. That is, current goal modeling languages have limited support for reasoning about changing beliefs and opinions, and their effect on the goal model. A stakeholder may change his or her opinion, but it is not always directly clear what its effect is on the goal model. Similarly, with existing goal modelling languages one can change a part of the goal model, but it is not possible to reason about whether or not this new goal model is consistent with the underlying beliefs and arguments. This becomes more problematic if the participants constructing the goal model change, since modeling decisions made by one group of stakeholders may conflict with the underlying beliefs put forward by another group of stakeholders.

To summarize, what is needed is a way of recording the rationales (beliefs, opinions, discussions, ideas) underlying a goal model. It should be possible to see how these rationales changed during the goal modeling process, and the rationales hould be clearly linked to the various elements of the resulting goal model. In order to be able to do this, we propose a framework with tool-support that combines traditional goal modeling approaches with argumentation techniques from Artificial Intelligence (AI) research~\cite{BenchCaponDunne2007}. We have identified \textbf{five important requirements} for our framework:

\begin{enumerate}
\item 
The argumentation techniques should be close to the actual discussions of stakeholders or designers in the early requirements engineering phase.
\item 
The framework must have formal traceability links between elements of the goal model and underlying arguments.
\item 
Using these traceability links, it must be possible to compute the effect of changes in the goal model on the underlying arguments, and vice versa.
\item 
There should be a methodology for the framework to guide the practitioners in its application in real cases.
\item 
The framework must have software tool support.
\end{enumerate}

Our framework, which we call \emph{RationalGRL}, combines the Goal-oriented Requirements Language (GRL)~\cite{Amyot:2010:EGM:1841349.1841356} with a technique from argumentation theory and discourse modeling called \emph{argument schemes} (or argumentation schemes~\cite{walton-etal2008}). Argument schemes are reusable patterns of reasoning that capture the typical ways in which humans argue and reason, and they are very well suited for modeling discussions about a goal model, as they can guide users in systematically deriving conclusions and making assumptions explicit~\cite{murukannaiah2015}. One argument scheme that is important when reasoning about goals is the argument scheme for practical reasoning~\cite{walton1990,atkinson2007}, which has been used for capturing multi-agent planning \cite{medellin2013planning}, dialogues about safety critical actions \cite{tolchinsky2012deliberation} and software design discussions~\cite{BlackEtal2013}.

Inspired by the work on practical reasoning from Artificial Intelligence, most notably Atkinson and Bench-Capon~\cite{atkinson2007}, we have developed a list of argument schemes that can be used to analyse and guide stakeholders' discussions about goal models. Our approach thus provides a rationalization to the elements of the goal model in terms of underlying arguments, and helps in understanding why parts of the model have been accepted and others have been rejected. Our list of argument schemes was constructed by performing an extensive case study in which we analyzed a set of transcripts containing more than 4 hours of discussions among designers of a traffic simulator information system. This ensures that the argumentation schemes we propose are close to actual real-world discussions stakeholders have (\textbf{requirement 1}).

The meta-model of the RationalGRL framework clearly specifies the traceability links between the arguments based on the schemes and the GRL models (\textbf{requirement 2}). In addition to this meta-model, we provide formal semantics for RationalGRL by formalising the GRL language in propositional logic and rendering arguments about a GRL model as a formal argumentation framework~\cite{Dung1995}. We then formally capture the link between argumentation and goal modelling as a set of algorithms for applying argument schemes and critical questions about goal models. These formal traceability links allow us to compute the effect of the arguments and counterarguments proposed in a discussion on a GRL model (\textbf{requirement 3}). In other words, we can determine whether the elements of a GRL model are acceptable given potentially contradictory opinions of stakeholders. Thus, we add a new formal evaluation technique for goal models that allows us to assess the \emph{acceptability} of elements of a goal model (in addition to their \emph{satisfiability}~\cite{Amyot:2010:EGM:1841349.1841356}).

Because we want RE practitioners in the field to be able to use our RationalGRL framework (\textbf{requirement 4}), we also propose a methodology for using RationalGRL, which consists of developing goal models and posing arguments based on schemes in an integrated way. To show that the RationalGRL methodology can be used in a real case, we illustrate the steps of our methodology with the traffic simulator case study. Finally, we have developed a web-based prototype\footnote{\url{insertURL}} for building goal models and arguing about them, which acts as a supporting tool to the RationalGRL methodology (\textbf{requirement 5}).

The rest of this article is organized as follows. Section~\ref{sect:background} introduces our running example, the Goal-oriented Requirements Language (GRL)~\cite{Amyot:2010:EGM:1841349.1841356}, the Practical Reasoning Argument Scheme (PRAS)~\cite{atkinson2007} and discusses some of our previous work on combining GRL and PRAS. Section~\ref{sect:overview} provides a brief and high-level overview of our framework, together with a metamodel and the methodology. Section~\ref{sect:gmas} contains an in depth explanation of how we obtained an initial set of argument schemes and critical questions by annotating transcripts from discussions about an information system, and in Section~\ref{sect:examples} we provide several examples of these schemes and questions. In Section~\ref{sect:formalframework} we provide formal semantics for GRL and show how argumentation semantics \cite{Dung1995} can be used to compute which arguments are accepted and which are rejected. We also develop various algorithms for the argument schemes in this section. In Section~\ref{sect:tool} we provide a brief overview of the prototype tool we developed for RationalGRL. Finally, Section~\ref{sect:discussion} contains a discussion, covering related work, future work, and a conclusion.
